---
title: "GEOG 5680 6680 00b Accessing CSBS services"
author: | 
  | Simon Brewer
  | Geography Department
  | University of Utah
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(png)
library(grid)
```

## Introduction

This class requires access to two pieces of software, R and RStudio. These are both open source and freely available, and I highly recommend installing them on your own computer (see the other document for installation instructions). If this is not possible, you can access these through the CSBS virtual lab. For this, you will need to access the [Horizon Client][horizID], and log on using your University credentials. You can access Horizon by downloading an app or through a web browser. A full description of how to do this can be found [here][csbsID]. 


Once you have connected, you can either open a desktop (the icon at the top) or open RStudio directly. I'd probably suggest working through the desktop, as this will make it easier to manage your file. Once you start this, look in the Apps menu for RStudio (a light blue circle with R in the middle), and click this. This should open a window that looks something like this:

```{r fig.width=6.5, fig.height=4., echo=FALSE}
img <- readPNG("images/rstudio_csbs.png")
grid.raster(img)
```

Please do not hesitate to contact me if you are having trouble accessing this. 

[csbsID]: https://support.csbs.utah.edu/get-help/self-support/virtual-lab.php

[horizID]: https://horizon.csbs.utah.edu