---
title: "GEOG 5680 Example Markdown"
author: "Simon Brewer"
date: "May 13, 2015"
output: html_document
---

# Iris dataset: plots and correlations 

## Introduction 
We are doing some exploratory statistics on the Fisher Iris dataset. Start by reading the file:
```{r}
iris = read.csv("iris.csv")
```

### Summary statistics
Calculate summary stats on all columns:
```{r}
summary(iris)
```

### Make a plot
Just a scatterplot:
```{r}
plot(iris$Sepal.Length, iris$Sepal.Width, 
     col=iris$Species, pch=16)
```
And a histogram:
```{r fig.keep='high'}
hist(iris$Sepal.Length, col="blue")
```

### Correlations
Start this with a list

- Correlation only the variables
- These are columns 1 to 4
```{r}
cor(iris[,1:4])
```

Be aware of [correlations][Correlation]!

### Discussion
Well, we have found out lots of things about the Iris. Flowers with wide petals tend to have long petals. Who knew? Well, we have found out lots of things about the Iris. Flowers with wide petals tend to have long petals. Who knew? Well, we have found out lots of things about the Iris. Flowers with wide petals tend to have long petals. Who knew? Well, we have found out lots of things about the Iris. Flowers with wide petals tend to have long petals. Who knew?

[Correlation]: http://xkcd.com/552/