---
title: "exercise11md"
author: "Daniel Kirkegaard"
date: "6/15/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# setwd("~/Documents/intro to R/exercise 11")
getwd()
list.files()
require(data.table)
```

## Read in the data using fread
```{r}
fread("Temperature.csv")
temp1 = fread("Temperature.csv")
```

## Extract all winter observations
```{r}
class(temp1)
winter = temp1[Season == "winter"]
```

## Extract all winter observations for zone NC
```{r}
winter = temp1[Season == "winter", Area == "NC"]
```

## Select only the columns Area, Season and Temperature
```{r}
AST = c("Area", "Temperature", "Season")
temp1[ , ..AST]
```

## Select only the columns Area and Temperature but only for winter observations
```{r}
WinterAT = c("Area", "Temperature")
winter[ , ..WinterAT]
```

## Find the total number of observations in winter
```{r}
dim(winter)
```

## Calculate the mean temperature and mean salinity in winter (Note that there are missing values so will have to use na.rm = TRUE)
```{r}
wintmean = temp1[Season == 'winter',
                .(meantemp = mean(Temperature, na.rm = TRUE),
                  meansal = mean(Salinity, na.rm = TRUE))]
wintmean
```

## Find the number of observations per station in winter
```{r}
table(winter$Station)
```

## Find the number of observations per station per season
```{r}
table(temp1$Station, temp1$Season)
```

## Estimate average temperatures by month
```{r}
meanmonthtemp = temp1[, .(meanmonthtemp = mean(Temperature, na.rm=TRUE)), by = Month]
meanmonthtemp
```

## Estimate average temperatures by month by area
```{r}
meanmonthareatemp = temp1[, .(meanmonthareatemp = mean(Temperature, na.rm=TRUE)), by = .(Area, Month)]
meanmonthareatemp
```

## Plot the output of the previous question using ggplot2 using the geom_line() geometry
```{r}
require(ggplot2)
library(ggplot2)
DKplot = ggplot(meanmonthareatemp, aes(x=Month, y=meanmonthareatemp, col= Area)) + ggtitle("Mean Monthly Temperature by Area") + scale_x_discrete(name = "Month", limits = c("1","2","3","4","5","6","7","8","9","10","11","12")) + scale_y_continuous(name = "Temperature")
DKplot + geom_line(aes (y=meanmonthareatemp))
```

